---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Merve Onal"
date: "Spring 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1 I set up the working directory and load the necessary packages
setwd('/home/guest/EDA_Spring2024/Data/Raw')

#First step is to check my working directory
getwd()

#install.packages
library(tidyverse)
library(agricolae)

# I import the data 
raw_data <- read.csv("NTL-LTER_Lake_ChemistryPhysics_Raw.csv")

# Convert the "sampledate" column to a date object
raw_data$sampledate <- as.Date(raw_data$sampledate, format = "%m/%d/%y")

# I check if sample date is a date object 
class(raw_data$sampledate)

#2
# Define my custom ggplot theme
library(ggplot2)
custom_theme <- theme_minimal() +
  theme(
    text = element_text(size = 12, color = "black"),     
    axis.title = element_text(face = "bold"),            
    axis.text = element_text(color = "blue"),           
    panel.background = element_rect(fill = "lightgray"),
    panel.grid.major = element_line(color = "gray"),      
    panel.grid.minor = element_blank()                    
  )


# I set my custom theme as the default theme
theme_set(custom_theme)



```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: There is no significant difference in mean lake temperature recorded during July across different depths in all lakes.
Ha: There is a significant difference in mean lake temperature recorded during July across different depths in all lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
# Load the dplyr and tidyr packages
library(dplyr)
library(tidyr)

# Here I filter for records with dates in July, select specific columns, and remove NAs
filtered_data <- raw_data %>%
  filter(format(sampledate, "%m") == "07") %>%  
  select(lakename, year4, daynum, depth, temperature_C) %>%  
  drop_na()  

# View the first few rows of the filtered dataset
head(filtered_data)

#5
# Load the ggplot2 package
library(ggplot2)

# Create a scatter plot with a smoothed line
scatter_plot <- ggplot(data = filtered_data, aes(x = depth, y = temperature_C)) +
  geom_point(color = "blue", size = 3, alpha = 0.5) +  # Scatter points
  geom_smooth(method = "lm", se = FALSE, color = "red") +  # Smoothed line (linear model)
  
  # Customize my plot's appearance
  labs(
    title = "Scatter Plot of Temperature by Depth",
    x = "Depth (m)",
    y = "Temperature (°C)"
  ) +
  theme_minimal() +  
  theme(
    panel.grid.major = element_blank(),  
    panel.grid.minor = element_blank(),  
    axis.title = element_text(face = "bold"),  
    axis.text = element_text(size = 12),      
    plot.title = element_text(size = 16, hjust = 0.5)  
  ) +
  ylim(0, 35)  

# Display the scatter plot
scatter_plot


```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The distribution of the data suggests that there may not be a purely linear relationship, rather, one that is more sigmodial.

7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
# Linear regression model
linear_model <- lm(temperature_C ~ depth, data = filtered_data)

# Display the summary of the linear regression model
summary(linear_model)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:The linear regression model shows that changes in depth play a significant role in the differences observed in lake temperature, accounting for nearly 74% of the variability. The model is based on 9,726 data points and indicates a relationship between temperature and depth that is statistically significant. The p-values for both the intercept and slope of the "depth" variable are almost zero, meaning they are unlikely to be zero. This suggests that for every 1-meter increase in depth, the temperature will decrease by approximately 1.946 degrees Celsius, as predicted by the estimated slope of the "depth" variable.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9

# Here I create a list of candidate models with different combinations of predictor variables
candidate_models <- list(
  model1 = lm(temperature_C ~ year4, data = filtered_data),
  model2 = lm(temperature_C ~ daynum, data = filtered_data),
  model3 = lm(temperature_C ~ depth, data = filtered_data),
  model4 = lm(temperature_C ~ year4 + daynum, data = filtered_data),
  model5 = lm(temperature_C ~ year4 + depth, data = filtered_data),
  model6 = lm(temperature_C ~ daynum + depth, data = filtered_data),
  model7 = lm(temperature_C ~ year4 + daynum + depth, data = filtered_data)
)

#Using 
model7 = lm(temperature_C ~ year4 + daynum + depth, data = filtered_data)
step(model7)


# Calculate AIC for each model
AIC_values <- sapply(candidate_models, AIC)

# Find the model with the lowest AIC
best_model <- names(AIC_values)[which.min(AIC_values)]

# Display AIC values and the best model
AIC_values
best_model

#10
# Fit a multiple regression model with year4, daynum, and depth as predictors
multiple_regression_model <- lm(temperature_C ~ year4 + daynum + depth, data = filtered_data)

# Display the summary of the multiple regression model
summary(multiple_regression_model)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC method suggests using three variables, year4, daynum, and depth, to predict temperature variability in the North Temperate Lakes LTER data. These variables provide the best predictors for explaining temperature changes, and together they explain 74.12% of the observed temperature variance. By comparing the model that uses all three variables with the model that uses only depth, we found that the model with all three variables provides a slight improvement in explaining temperature variance. This suggests that year4 and daynum contribute additional information to the prediction of temperature beyond what can be explained by depth alone. The model including all three variables is better at explaining temperature variance than the model with only depth, which explains 73.87% of the observed temperature variance.




---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
# Now I run an ANOVA test to compare mean temperatures among lakes
anova_model <- aov(temperature_C ~ lakename, data = filtered_data)

# Display the summary of the ANOVA model
summary(anova_model)

# Fit a linear model with lakename as a predictor
linear_model_lakes <- lm(temperature_C ~ lakename, data = filtered_data)

# Display the summary of the linear model
summary(linear_model_lakes)


```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, there is a significant difference in mean temperature among the lakes in the dataset for the month of July. Both the ANOVA test and the linear model have provided strong evidence of this significant difference. The ANOVA results showed that the p-value associated with the F-statistic is very close to zero (<2e-16), which is highly significant. This indicates that at least one lake has a different mean temperature from the others. Similarly, the linear model results indicated that the p-values for each lake (compared to the reference lake) are all very close to zero, indicating that the differences in mean temperatures for each lake are statistically significant.



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
# Create the scatter plot with separate colors for each lake
ggplot(filtered_data, aes(x = depth, y = temperature_C, color = lakename)) +
  geom_point(alpha = 0.5) +  
  geom_smooth(method = "lm", se = FALSE) +  
  ylim(0, 35) + 
  labs(
    x = "Depth (meters)",
    y = "Temperature (°C)",
    title = "Temperature by Depth for Different Lakes in July",
    color = "Lake Name"
  ) +
  theme_minimal() 

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
# Load stats package
library(stats)

# Perform Tukey's HSD test to compare means of different lakes
tukey_result <- HSD.test(aov(temperature_C ~ lakename, data = filtered_data),"lakename", group = T) 

# Print the results
print(tukey_result)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer:The results of the statistical analysis show that the temperature of Peter Lake is significantly different from Crampton Lake, East Long Lake, Hummingbird Lake, Tuesday Lake and West Long Lake. The adjusted p-value for the comparison between Peter Lake and Crampton Lake is less than 0.05, indicating a statistically significant difference in their means. Similarly, the adjusted p-value for the comparison between Peter Lake and Hummingbird Lake is also less than 0.05, suggesting a statistically significant difference in means. However, there is no statistically significant difference in means between Peter Lake and Paul Lake or between Peter Lake and Ward Lake. The adjusted p-value for the comparison between Peter Lake and Paul Lake is greater than 0.05. Likewise, the adjusted p-value for the comparison between Peter Lake and Ward Lake is also greater than 0.05.
Consequently based on this information, if any lake has a statiscally distinct mean from all other lakes we can say it is not statistically distinct from all the other lakes. Answer is no.


17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: If we were to specifically compare the mean temperatures of Peter Lake and Paul Lake and determine whether they have distinct mean temperatures, we could perform an independent samples t-test.



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
# Subset the data for Crampton Lake and Ward Lake
crampton_lake_data <- filtered_data[filtered_data$lakename == "Crampton Lake", "temperature_C"]
ward_lake_data <- filtered_data[filtered_data$lakename == "Ward Lake", "temperature_C"]

# Perform a two-sample t-test
t_test_result <- t.test(crampton_lake_data, ward_lake_data)

# Print the t-test result
print(t_test_result)
```

>Answer: The Welch Two-Sample t-test results indicate the following:
1. The test does not provide enough evidence to conclude that the mean temperatures for Crampton Lake and Ward Lake in July are significantly different.
2. The null hypothesis that the mean temperatures of Crampton Lake and Ward Lake are equal cannot be rejected based on the t-test.
This result does not match the answer for part 16, which suggested that there were statistically significant differences in mean temperatures between lakes based on Tukey's HSD test. In part 16, it was concluded that Peter Lake had the same mean temperature, statistically speaking, as Paul Lake and Ward Lake. However, the t-test result for Crampton Lake and Ward Lake in part 18 suggests that their mean temperatures are not statistically distinct.
The difference in results between part 16 and part 18 may be due to the different statistical tests used and their underlying assumptions. Tukey's HSD test considers multiple pairwise comparisons among lakes, while the t-test focuses on a specific comparison between two lakes. Additionally, the t-test used a Welch correction to account for unequal variances, which can affect the results. These differences highlight the importance of choosing appropriate statistical tests and considering the specific research question and data characteristics.

###Crampton Lake and Ward Lake in the same group indicating they have statistically similar means.
