---
title: "Assignment 8: Time Series Analysis"
author: "Merve Onal"
date: "Spring 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A08_TimeSeries.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

## Set up 
1. Set up your session:

* Check your working directory
* Load the tidyverse, lubridate, zoo, and trend packages
* Set your ggplot theme

```{r}
getwd()
#install.packages("tidyverse")
library(tidyverse)
library(lubridate)
#install.packages("zoo")
library(zoo)
#install.packages("trend")
library(trend)

# Here I set the ggplot theme
theme_set(theme_minimal())

```


2. Import the ten datasets from the Ozone_TimeSeries folder in the Raw data folder. These contain ozone concentrations at Garinger High School in North Carolina from 2010-2019 (the EPA air database only allows downloads for one year at a time). Import these either individually or in bulk and then combine them into a single dataframe named `GaringerOzone` of 3589 observation and 20 variables. 

```{r, message = FALSE}
#1

setwd('/home/guest/EDA_Spring2024/Data/Raw/Ozone_TimeSeries')

# Here is the list of file names to import
file_names <- c(
  "EPAair_O3_GaringerNC2010_raw.csv",
  "EPAair_O3_GaringerNC2011_raw.csv",
  "EPAair_O3_GaringerNC2012_raw.csv",
  "EPAair_O3_GaringerNC2013_raw.csv",
  "EPAair_O3_GaringerNC2014_raw.csv",
  "EPAair_O3_GaringerNC2015_raw.csv",
  "EPAair_O3_GaringerNC2016_raw.csv",
  "EPAair_O3_GaringerNC2017_raw.csv",
  "EPAair_O3_GaringerNC2018_raw.csv",
  "EPAair_O3_GaringerNC2019_raw.csv"
)

# Importing and combining the datasets
GaringerOzone <- bind_rows(
  lapply(file_names, function(file) read.csv(file, header = TRUE))
)

# Here I check the dimensions of the combined dataframe
dim(GaringerOzone)

```

## Wrangle

3. Set your date column as a date class.

4. Wrangle your dataset so that it only contains the columns Date, Daily.Max.8.hour.Ozone.Concentration, and DAILY_AQI_VALUE. 

5. Notice there are a few days in each year that are missing ozone concentrations. We want to generate a daily dataset, so we will need to fill in any missing days with NA. Create a new data frame that contains a sequence of dates from 2010-01-01 to 2019-12-31 (hint: `as.data.frame(seq())`). Call this new data frame Days. Rename the column name in Days to "Date".

6. Use a `left_join` to combine the data frames. Specify the correct order of data frames within this function so that the final dimensions are 3652 rows and 3 columns. Call your combined data frame GaringerOzone. 

```{r}
# 3
GaringerOzone$Date <- as.Date(GaringerOzone$Date, format = "%m/%d/%Y")

class(GaringerOzone$Date)

# 4

GaringerOzone <- GaringerOzone %>%
  select(Date, Daily.Max.8.hour.Ozone.Concentration, DAILY_AQI_VALUE)

str(GaringerOzone)

# 5

# Here I create a sequence of dates from 2010-01-01 to 2019-12-31
start_date <- as.Date("2010-01-01")
end_date <- as.Date("2019-12-31")
Days <- as.data.frame(seq(from = start_date, to = end_date, by = "day"))

# Here I create a "Days" dataframe with a complete sequence of dates
start_date <- as.Date("2010-01-01")
end_date <- as.Date("2019-12-31")
Days <- as.data.frame(seq(from = start_date, to = end_date, by = "day"))
colnames(Days) <- "Date"

#  Fill in missing days
GaringerOzone <- Days %>%
  left_join(GaringerOzone, by = "Date")


# Rename the column to "Date"
colnames(Days) <- "Date"


# 6

# Combining the data frames
GaringerOzone <- left_join(Days, GaringerOzone, by = "Date")

# Checking the dimensions of the combined data frame
dim(GaringerOzone)

```

## Visualize

7. Create a line plot depicting ozone concentrations over time. In this case, we will plot actual concentrations in ppm, not AQI values. Format your axes accordingly. Add a smoothed line showing any linear trend of your data. Does your plot suggest a trend in ozone concentration over time?

```{r}
#7

# Filtering out rows with missing ozone concentration values (NA)
filtered_data <- GaringerOzone %>%
  filter(!is.na(Daily.Max.8.hour.Ozone.Concentration))

# Checking if there are any valid data points left
if (nrow(filtered_data) > 0) {

  # Creating a line plot with ggplot2
  ggplot(filtered_data, aes(x = Date, y = Daily.Max.8.hour.Ozone.Concentration)) +
    geom_line() + 
    geom_smooth(method = "lm", se = FALSE, color = "red") +  
    labs(
      title = "Ozone Concentrations Over Time",
      x = "Date",
      y = "Ozone Concentration (ppm)"
    ) +
    scale_x_date(date_labels = "%Y", date_breaks = "1 year") +  
    theme_minimal()
} else {
  cat("No valid data points for the plot.")
}

print(plot)



```

>Answer: From the plot, we notice a slightly decreasing trend throughout the 10 year period from 2010 to 2020. Furthermore, we also can notice seasonality changes and fluctuations within each of the given years. For example, we notice peaks on specific months, and troughs on the others. 

## Time Series Analysis

Study question: Have ozone concentrations changed over the 2010s at this station? 

8. Use a linear interpolation to fill in missing daily data for ozone concentration. Why didn't we use a piecewise constant or spline interpolation?

```{r}
#8
 GaringerOzone$Daily.Max.8.hour.Ozone.Concentration <- 
  na.approx(GaringerOzone$Daily.Max.8.hour.Ozone.Concentration, na.rm = FALSE)

```

> Answer: Linear interpolation was chosen over piecewise constant to fill in the missing ozone data because it accounts for the steady rate of change between data points, which aligns with the nature of ozone fluctuations. Piecewise constant interpolation does not account for daily variations. Spline interpolation best fit for quadratic data and could introduce artificial swings and potentially misrepresent the true trend.


9. Create a new data frame called `GaringerOzone.monthly` that contains aggregated data: mean ozone concentrations for each month. In your pipe, you will need to first add columns for year and month to form the groupings. In a separate line of code, create a new Date column with each month-year combination being set as the first day of the month (this is for graphing purposes only)

```{r}
#9


# Add columns for year and month
GaringerOzone <- GaringerOzone %>%
  mutate(year = year(Date),
         month = month(Date))

# Aggregate data: mean ozone concentrations for each month
GaringerOzone.monthly <- GaringerOzone %>%
  group_by(year, month) %>%
  summarize(mean_ozone = mean(Daily.Max.8.hour.Ozone.Concentration))

# Create a new Date column with each month-year combination set as the first day of the month
GaringerOzone.monthly <- GaringerOzone.monthly %>%
  mutate(Date = as.Date(paste(year, month, "01", sep = "-")))

# View the first few rows of the aggregated data
head(GaringerOzone.monthly)


```


10. Generate two time series objects. Name the first `GaringerOzone.daily.ts` and base it on the dataframe of daily observations. Name the second `GaringerOzone.monthly.ts` and base it on the monthly average ozone values. Be sure that each specifies the correct start and end dates and the frequency of the time series.

```{r}
#10
# Daily Time Series
start_year_daily <- year(first(GaringerOzone$Date))
monthfirst_daily <- month(first(GaringerOzone$Date))
GaringerOzone.daily.ts <- ts(GaringerOzone$Daily.Max.8.hour.Ozone.Concentration, 
                             start = c(start_year_daily, monthfirst_daily, 1), 
                             frequency = 365)

# Monthly Time Series
start_year_monthly <- year(first(GaringerOzone.monthly$Date))
monthfirst_monthly <- month(first(GaringerOzone.monthly$Date))
GaringerOzone.monthly.ts <- ts(GaringerOzone.monthly$mean_ozone, 
                               start = c(2010, 1), 
                               frequency = 12)

```


11. Decompose the daily and the monthly time series objects and plot the components using the `plot()` function.

```{r}
#11
# Decomposing the daily time series
GaringerOzone.daily.decomp <- stl(GaringerOzone.daily.ts, s.window = "periodic")
# Plotting the components
plot(GaringerOzone.daily.decomp)
# Decomposing the monthly time series
GaringerOzone.monthly.decomp <- stl(GaringerOzone.monthly.ts, s.window = "periodic")
# Plotting the components
plot(GaringerOzone.monthly.decomp)


```


12. Run a monotonic trend analysis for the monthly Ozone series. In this case the seasonal Mann-Kendall is most appropriate; why is this?

```{r}
#12

#install.packages("trend")

library(trend)
library(Kendall)


# Running the Seasonal Mann-Kendall test
result <- Kendall::SeasonalMannKendall(GaringerOzone.monthly.ts)

print(result)


```

> Answer: The Seasonal Mann-Kendall test is best suited for analyzing trends in monthly ozone levels. This test takes into account seasonal variations caused by various factors such as (temperature etc). Using a test that can handle these seasonal changes is important to accurately identify trends over time. The Seasonal Mann-Kendall test is designed to address these seasonal cycles, which can otherwise mask or distort the true underlying trend in the data when using a standard Mann-Kendall test. Based on my test results, there is a statistically significant decreasing trend in the monthly Ozone series at a 5% significance level with a z-value of -1.963 and a p-value of 0.04965.

13. Create a plot depicting mean monthly ozone concentrations over time, with both a geom_point and a geom_line layer. Edit your axis labels accordingly.

```{r}
# 13

library(ggplot2)

plot <- ggplot(GaringerOzone.monthly, aes(x = Date, y = mean_ozone)) +
  geom_point() +  
  geom_line() +   
  labs(
    title = "Mean Monthly Ozone Concentrations Over Time",
    x = "Time",
    y = "Mean Ozone Concentration (ppm)"
  ) +
  theme_minimal()

print(plot)



```

14. To accompany your graph, summarize your results in context of the research question. Include output from the statistical test in parentheses at the end of your sentence. Feel free to use multiple sentences in your interpretation.

> Answer: Based on the analysis of monthly mean ozone concentration data from July 2010 to February 2019 using the Seasonal Mann-Kendall trend test, it appears that there has been a consistent but slight decrease in ozone concentration each month when accounting for seasonal variation. The results suggest a statistically significant decreasing trend in ozone levels over this period (z = -1.963, p-value = 0.04965), but it is important to note that the detected trend is at the margin of statistical significance given the p-value being close to the conventional threshold of 0.05. To confirm this trend and rule out the possibility of random fluctuations or to investigate potential underlying factors contributing to this change, further data and analysis might be needed.


15. Subtract the seasonal component from the `GaringerOzone.monthly.ts`. Hint: Look at how we extracted the series components for the EnoDischarge on the lesson Rmd file.

16. Run the Mann Kendall test on the non-seasonal Ozone monthly series. Compare the results with the ones obtained with the Seasonal Mann Kendall on the complete series.

```{r}
#15

# Extracting seasonal component from the decomposition results

GaringerOzone.monthly.components  <- 
  as.data.frame(GaringerOzone.monthly.decomp$time.series[,1:3])



# Subtracting the seasonal component from the original monthly time series
GaringerOzone.monthly.ts_deseasonalized <- 
  GaringerOzone.monthly.ts - GaringerOzone.monthly.components$seasonal

# Running the Mann-Kendall test on the deseasonalized time series
result_deseasonalized <- MannKendall(GaringerOzone.monthly.ts_deseasonalized)

print(result_deseasonalized)

#16
# Running the Mann-Kendall test on the deseasonalized data
mk_result <- Kendall::MannKendall(GaringerOzone.monthly.ts_deseasonalized)

print(mk_result)




```

> Answer: 
result_deseasonalized:
tau = -0.165, 2-sided pvalue =0.0075402
SMK:
tau = -0.143, 2-sided pvalue =0.046724
There is little evidence against null hypothesis, these results indicate a statistically significant decreasing trend in the deseasonalized ozone data and original data (p-value < 0.05). Both results are p-value < 0.05 which shows there is a decreasing trend in ozone. When remove the seasonality from the original data, there is a stronger downward trend. The non-seasonal p value is even stronger much more closer to zero, which shows when seasonality is removed, the trend is even stronger.