---
title: "Assignment 2: Coding Basics"
author: "Merve Onal"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics, Part 1

1.  Generate a sequence of numbers from one to 30, increasing by threes. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r basics 1}
threes_sequence <- seq(1, 30, by = 3) 
#1. here I inserted the command to generate the sequence as requested by the first question 

mean_value <- mean(threes_sequence)
#2. here I computed the mean of the sequence 

median_value <- median(threes_sequence) 
#3. here I computed the median of the sequence 

mean_greater_than_median <- mean_value > median_value
# here I compared the mean and the median
cat("Mean:", mean_value, "\n")
cat("Median:", median_value, "\n")
cat("Is mean greater than median?", mean_greater_than_median, "\n")

```
## Is mean greater than median? They are equal. FALSE

## Basics, Part 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r basics 2}
# In Vector 1 I have assigned the names of the students
student_names <- c("Adam", "Bella", "Charlie", "Daisy")

# In Vector 2 I assigned the test scores out of 100 points
test_scores <- c(75, 60, 90, 40)

# In Vector 3 I set Whether they passed the test (TRUE or FALSE) with a passing grade of 50
pass_status <- test_scores >= 50

# Now I combine the vectors into a data frame and by giving it a name
student_data <- data.frame(Name = student_names, Score = test_scores, Passed = pass_status)

# I specify the command to the data frame
student_data
```
9.  QUESTION: How is this data frame different from a matrix?

> Answer:Data frames are more versatile for diverse data, while matrices are more specialized for numerical data. A data frame allows us to have columns with different data types. A matrix just lets us be limited to a single data type for all elements. Furthermore, matrices so not have column names, making it not clear what each column represents. Matrices are less flexible in that every value in a matrix must be of the same data type.

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r basics 3}
# Here I define the function to check if a test score is passing or not by using the if else
check_passing_score <- function(score) {
if (score >= 50) {
return(TRUE)
} else {
return(FALSE)
}
}
# I apply the function to each test score in the vector and it worked
cat("Passing Grades:\n")


## Passing Grades:
for (score in test_scores) {
cat("Score:", score, "Passing:", check_passing_score(score), "\n")
}

# Now we define a function to check if a test score is passing using ifelse
check_and_print_passing_score_ifelse <- function(scores) {
result <- ifelse(scores >= 50, "TRUE", "FALSE")
cat(result, "\n")
}

# This is the vector of test scores
test_scores <- c(75, 60, 90, 40)

# Here I apply the function to the vector of test scores
cat("Passing Grades:\n")

# Now we define a function to check if a test score is passing using ifelse
check_and_print_passing_score_ifelse <- function(scores) {
result <- ifelse(scores >= 50, "TRUE", "FALSE")
cat(result, "\n")
}

# This is the vector of test scores
test_scores <- c(75, 60, 90, 40)

# Here I apply the function to the vector of test scores
cat("Passing Grades:\n")

## Passing Grades:
check_and_print_passing_score_ifelse(test_scores)

options(knitr.duplicate.label = "allow")
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: Both of if and else and ifelse options are worked.
